---
title: "Shirin_review"
author: "Shirin Parvin"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hi Faizo,

Great job attempting the assignment! 

Just a few suggestions before I continue with my review of the code. 

a. In the first chunk, you have the `dev.off()` to close the current graphics device. If I execute your .Rmd file using "Run all", this line of code throws an error ***"Error in dev.off() : cannot shut down device 1 (the null device)"***. I do not think including that line of code is crucial for your script. So, perhaps you can remove it to make your code run smoothly.

b. On line 17 of your code you had 4 "`" symbols while all the rest of your code chunks had 3 "`" symbols. Because of this, the "Run all" would not work. Please make sure all your chunks open and close with the same number of "`" symbols. Although this is a minor thing, it is easy to miss this and it can break your code. :)


Here are some of my suggestions as part of the review, which may help improve your assignment:

1. Instead of setting up the path to our directory, it might be worth it to upload the `fang_et_al_genotype.txt` and `snp_position.txt` files in your repository, so that your code is not dependent on the person actually having the data files. Just a good practise for coding to have all required files being in the repo.

2. I believe you do not need the following line of code since the snp_position dataframe is already arranged by SNP_ID.

```{r}
snp_position_arranged <- arrange(snp_position_Chr, SNP_ID)
```

3. In **"# Data Transposition"** part, it might be a good idea to convert the data into a dataframe using `data.frame()` after using transpose function `t()`, since `t()` converts the dataframe into a matrix, which might later arise issues in using it as a dataframe. Just good practise suggestion.

4. During generating the multiple files, as mentioned in the hint, it might be better to use `lapply()`. If you choose to do so, you can define the arranging and writing parts as a function first and then use `lapply()`. Just makes the code run faster.

5. In the Data Visualisation part, for "# Missing data and amount of homozygous,heterozygous and missing" section, you might need to replot it, since the question expects the plot to be a stacked plot, and for each individual sample and not overall.

